---
title: "Day 6: Custom Customs"
author: Jonas Nockert (@lemonad)
output:
  github_document:
    fig_width: 10
    fig_height: 5
    dev: png
always_allow_html: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(scipen = 999)
library(tidyverse)
library(ggplot2)
library(ggforce)
library(gridExtra)
library(grid)
library(kableExtra)
library(pryr)
library(rvest)
library(skimr)
library(stringr)
library(tibble)
```

I'm using R for this year's Advent of Code. Learning as I go, which is quite obvious here : )

It seems the biggest problem of part one is the grouping with empty lines. Especially since I struggle with parsing in R, it always seem to result in much more code than I initially expect.

```{r}
lines <- read_file("input/december06.input")
lines %>%
  str_split("\\n\\n") %>%
  unlist() %>%
  lapply(function(x) {
    str_replace_all(x, "\\n", "") %>%
    str_split("") %>%
    unlist() %>%
    unique %>%
    length
  }) %>%
  reduce(sum)
```

Okay, that's quite readable but, still, it's basically trial and error with the `unlist`'s.


## Part two
Uh oh, there's not much to reuse from part one here.

```{r}
lines %>%
  str_split("\\n\\n") %>%
  unlist() %>%
  str_split("\\n") %>%
  lapply(
    function(x) {
      str_split(x, "") %>%
      reduce(intersect) %>%
      length()
    }
  ) %>%
  unlist() %>%
  sum()
```

Hm, not the right answer (but someone else's answer so I'm assuming it is pretty close). Let's investigate...

```{r}
lines %>%
  str_split("\\n\\n") %>%
  unlist() %>%
  str_split("\\n") %>%
  tail(n = 2)
```
Aha, so the intersection of the last set of strings is "", I guess there is some new line at the end.

```{r}
lines %>%
  str_split("\\n\\n") %>%
  unlist() %>%
  .[str_detect(., "\\n$")]
```

Yes, there's only one trailing newline and it's on the very last line. Gah, I've forgotten to trim whitespace when reading the input. Let's try this again:

```{r}
lines <- trimws(read_file("input/december06.input"))
lines %>%
  str_split("\\n\\n") %>%
  unlist() %>%
  str_split("\\n") %>%
  lapply(
    function(x) {
      str_split(x, "") %>%
      reduce(intersect) %>%
      length()
    }
  ) %>%
  unlist() %>%
  sum()
```

That did the trick : )
