---
title: "Day 3: Toboggan Trajectory"
author: Jonas Nockert (@lemonad)
output:
  github_document:
    fig_width: 10
    fig_height: 5
    dev: png
    df_print: kable
always_allow_html: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(scipen = 999)
library(tidyverse)
library(ggplot2)
library(ggforce)
library(gridExtra)
library(grid)
library(kableExtra)
library(pryr)
library(rvest)
library(skimr)
library(stringr)
library(tibble)
```

I'm using R for this year's Advent of Code. Learning as I go, which is quite obvious here : )


Testing with the example data seems like a good first step. Not sure what representation is most useful in R, perhaps matrix format?

```{r}
example_input <- trimws(
"
..##.......
#...#...#..
.#....#..#.
..#.#...#.#
.#...##..#.
..#.##.....
.#.#.#....#
.#........#
#.##...#...
#...##....#
.#..#...#.#
"
)

example_slope <- str_replace_all(example_input, "\\n", "") %>%
  str_split("") %>%
  unlist() %>%
  matrix(nrow = 11, ncol = 11, byrow = TRUE)
```


Let's plot the resulting slope matrix to see that it looks correct (which turned out easier said than done):

```{r}
plot_slope <- function(slope, title) {
  slope %>%
    as_tibble() %>%
    mutate(y = row_number()) %>%
    pivot_longer(cols = 1:11, names_to = "x") %>%
    transmute(
      y = y,
      x = as.integer(str_sub(x, 2)),
      value = as.integer(ifelse(value == "#", 1, 0))
    ) %>%
    ggplot(aes(x = x, y = -y, fill = value)) +
    geom_tile() +
    scale_fill_gradient(low = "white", high = "forestgreen") +
    coord_equal() +
    xlab("") +
    ylab("") +
    labs(title = title) +
    theme_void() +
    theme(legend.position = "none")
}

plot_slope(example_slope, "Example slope")
```

Using the example input, the expectation is that we encounter 7 trees.

```{r}
ride_slope <- function(slope, step) {
  n_rows <- dim(slope)[1]
  n_cols <- dim(slope)[2]
  pos <- c(1, 1)

  n_trees <- 0
  while (pos[1] <= n_rows) {
    if (slope[pos[1], pos[2]] == "#") {
      n_trees <- n_trees + 1
    }
    pos <- c(pos[1] + step[1], 1 + (pos[2] + step[2] - 1) %% n_cols)
  }
  return(n_trees)
}

n_trees <- ride_slope(example_slope, c(1, 3))
print(
  paste("We encountered", n_trees, "trees.")
)
```

Sounds about right so let's try it on the real input:

```{r}
lines <- read_lines("./input/december03.input")
rows <- length(lines)
slope <- str_replace_all(lines, "\\n", "") %>%
  str_split("") %>%
  unlist() %>%
  matrix(nrow = rows, byrow = TRUE)

plot_slope(slope, "Slope")
```

```{r}
n_trees <- ride_slope(slope, c(1, 3))
print(
  paste("We encountered", n_trees, "trees.")
)
```

## Part 2
For the second part, we will try a few different traverses. Here we expect the product of encountered trees to be 336.

```{r}
prod_trees <- function(slope) {
  return(
    prod(
      ride_slope(slope, c(1, 1)),
      ride_slope(slope, c(1, 3)),
      ride_slope(slope, c(1, 5)),
      ride_slope(slope, c(1, 7)),
      ride_slope(slope, c(2, 1))
    )
  )
}

print(
  paste("Product of trees encountered:", prod_trees(example_slope))
)
```

Finally, on the slope we've been given:

```{r}
print(
  paste("Product of trees encountered:", prod_trees(slope))
)
```
