---
title: "Day 6: Handy Haversacks"
author: Jonas Nockert (@lemonad)
output:
  github_document:
    fig_width: 10
    fig_height: 5
    dev: png
always_allow_html: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(scipen = 999)
library(tidyverse)
library(ggplot2)
library(ggforce)
library(gridExtra)
library(grid)
library(kableExtra)
library(pryr)
library(rvest)
library(skimr)
library(stringr)
library(tibble)
```

I'm using R for this year's Advent of Code. Learning as I go, which is quite obvious here : )

It seems like a good idea to start with the given example and work our way up
to the real input. It looks like splitting on the word "contain" and removing
all occurrences of the words "bag" and "bags" is a good start.
[Eventually, I decided to also remove "no other" as well.]

```{r}
example_input <- trimws("
light red bags contain 1 bright white bag, 2 muted yellow bags.
dark orange bags contain 3 bright white bags, 4 muted yellow bags.
bright white bags contain 1 shiny gold bag.
muted yellow bags contain 2 shiny gold bags, 9 faded blue bags.
shiny gold bags contain 1 dark olive bag, 2 vibrant plum bags.
dark olive bags contain 3 faded blue bags, 4 dotted black bags.
vibrant plum bags contain 5 faded blue bags, 6 dotted black bags.
faded blue bags contain no other bags.
dotted black bags contain no other bags.
")

parse_bags <- function(str_or_path) {
  lines <- read_lines(str_or_path)
  bags <- lines %>%
    str_replace_all("(no other bags|bags|bag)", "") %>%
    str_split("contain") %>%
    lapply(trimws, whitespace = "[ \n\r\t\\.]")
}

raw_bags <- parse_bags(example_input)
raw_bags %>%
  head(3) %>%
  kable(
    caption = "Examples of bags and their content",
    col.names = NULL,
    format = "html"
  )
```

Not sure if it's strictly necessary but let's separate the list
into outer bags and inner bags:

```{r}
separate_bags <- function(bags) {
  outer_bags <- bags %>%
    map(1) %>%
    unlist()
  inner_bags <- bags %>%
    map(2)
  list(outer = outer_bags, inner = inner_bags)
}

bags <- separate_bags(raw_bags)
bags$outer %>% head(1)
bags$inner %>% head(1)
```

So, can we find the bags that hold our _shiny gold_ bag directly using the
data structures we've set up?

```{r}
holds_directly <- bags$outer[
  str_detect(bags$inner, "shiny gold")
] %>%
  unlist() %>%
  unique()
holds_directly
```

Yes, that looks like the correct answer. Now, it looks like we should
do the same thing for the bag colors contained by those bag colors, etc.

```{r}
find_bag_options <- function(color, all_bags) {
  inner_func <- function(colors_to_look_for) {
    matching_bags <- all_bags$inner %>%
      str_split(",") %>%
      lapply(
        function(x) {
          trimws(x, whitespace = "[ \\d]")
        }
      ) %>%
      lapply(
        function(x) {
          x %in% colors_to_look_for
        }
      ) %>%
      lapply(any) %>%
      unlist()
    all_bags$outer[matching_bags]
  }

  # In retrospective, this should have been a recursive solution!
  all_colors <- vector()
  inner <- color
  while (!is_empty(inner)) {
    inner <- inner_func(inner) %>%
      setdiff(all_colors)
    all_colors <- union(inner, all_colors)
  }
  all_colors
}

find_bag_options("shiny gold", bags)
```

Four (4), like we expected. I think this solution could be refactored to a much nicer one but let's try it on the real input:

```{r}
raw_bags <- parse_bags("input/december07.input")
bags <- separate_bags(raw_bags)
find_bag_options("shiny gold", bags) %>%
  length()
```

Correct! I might go back and fix things here but, for now, onwards to part two!

## Part two
Again, we'll start with the given example, reusing the functions from part one:

```{r}
example_input <- trimws("
shiny gold bags contain 2 dark red bags.
dark red bags contain 2 dark orange bags.
dark orange bags contain 2 dark yellow bags.
dark yellow bags contain 2 dark green bags.
dark green bags contain 2 dark blue bags.
dark blue bags contain 2 dark violet bags.
dark violet bags contain no other bags.
")
raw_bags <- parse_bags(example_input)
bags <- separate_bags(raw_bags)
```

It looks like this forms a tree structure with our shiny gold bag at the top, followed by 2 dark red bags, etc. This is something we could recurse down, although it feels like the mention of "topologically impractical" hints toward a cyclical graph (which is a recurring theme in previous AoCs). Let's see:

```{r}
recurse <- function(color, all_bags, acc) {
  bag_strings <- bags$inner[bags$outer == color] %>%
    str_split(",") %>%
    unlist() %>%
    trimws()

  # Handle "no other bags", i.e. c("") after parsing.
  if (length(bag_strings) == 1 && str_length(bag_strings) == 0) {
    return(acc)
  }

  bag_counts <- bag_strings %>%
    str_extract("\\d+") %>%
    as.integer()

  bag_colors <- bag_strings %>%
    str_extract("[^\\d]+") %>%
    unlist() %>%
    trimws()

  s <- sum(
    mapply(
      function(x, y) {
        # Two bags, each containing three bags is 2 + 2 * 3 = 8 bags
        # in total.
        x + x * recurse(y, all_bags, 0)
      },
      bag_counts,
      bag_colors,
      USE.NAMES = TRUE
    )
  )
  acc + s
}

recurse("shiny gold", bags, 0)
```

Very good. We'll try it on the actual input:

```{r}
raw_bags <- parse_bags("input/december07.input")
bags <- separate_bags(raw_bags)
recurse("shiny gold", bags, 0)
```

Correct :)
